---
title: "Introduction to PopED"
author: "Andrew Hooker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Introduction to PopED}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

PopED computes optimal experimental designs for both population and individual studies based on nonlinear mixed-effect models. Often this is based on a computation of the Fisher Information Matrix (FIM).

To get started you need to define

* A model.
* An initial design (and design space if you want to optimize)
* The tasks to perform.

There are a number of functions to help you with these tasks.  See `?poped` for more information.  
 
Below is an example to introduce the package.  There are several other examples, as r-scripts, in the "examples" folder in the 
PopED installation directory located at:


```
system.file("examples", package="PopED")
```

## Define a model

Here we define a one-compartment pharmacokinetic model with linear absorption using an analytical solution.  In this case the solution is applicable for both multiple and single dosing.  Note that this function is also predefined in PopED as `ff.PK.1.comp.oral.md.CL`.

```{r,eval=TRUE}
library(Rcpp)
library(deSolve)
library(PopED)
```


```{r,include = FALSE}
set.seed(1234)
knitr::opts_chunk$set(cache = FALSE)
```



```{r struct_model}
ff <- function(model_switch,xt,parameters,poped.db){
  with(as.list(parameters),{
    N = floor(xt/TAU)+1
    y=(DOSE*Favail/V)*(KA/(KA - CL/V)) * 
      (exp(-CL/V * (xt - (N - 1) * TAU)) * (1 - exp(-N * CL/V * TAU))/(1 - exp(-CL/V * TAU)) - 
         exp(-KA * (xt - (N - 1) * TAU)) * (1 - exp(-N * KA * TAU))/(1 - exp(-KA * TAU)))  
    return(list( y=y,poped.db=poped.db))
  })
}
```

Next we define the parameters of this function, in this case log-normal distributions around the parameters (not `Favail`).  `DOSE` and `TAU` are defined as a covariates (`a`) so that we can optimize these values later.
```{r}
sfg <- function(x,a,bpop,b,bocc){
  parameters=c( V=bpop[1]*exp(b[1]),
                KA=bpop[2]*exp(b[2]),
                CL=bpop[3]*exp(b[3]),
                Favail=bpop[4],
                DOSE=a[1],
                TAU=a[2])
}
```

Now we define the residual unexplained variability (RUV) function, in this case the RUV has both an additive and proportional component.
```{r}
feps <- function(model_switch,xt,parameters,epsi,poped.db){
  returnArgs <- ff(model_switch,xt,parameters,poped.db) 
  y <- returnArgs[[1]]
  poped.db <- returnArgs[[2]]
 
  y = y*(1+epsi[,1])+epsi[,2]
  
  return(list(y=y,poped.db=poped.db)) 
}
```


We have now defined all aspects of the model.

## Define an initial design and design space

Now we define the model parameter values, the initial design  and design space for optimization.

In this example, the parameter values are defined for the fixed effects (`bpop`), the between-subject variability variances (`d`) and the residual variability variances (`sigma`).  We also fix the parameter `Favail` using `notfixed_bpop`, since we have only oral dosing and the parameter is not identifiable.  

Fixing a parameter means that we assume the parameter will not be estimated.  In addition, we fix the small additive RUV term, as this term is reflecting the higher error expected at low concentration measurements (limit of quantification measurements) and would typically be calculated from analytical assay methods (for example, the standard deviation of the parameter might be 20% of the limit of quantification).


For the initial design, we define two groups (`m=2`) of 20 individuals (`groupsize=20`), with doses of 20 mg or 40 mg every 24 hours (`a`).   The initial design has 5 sample times per individual (`xt`). 

For the design space, which can be searched during optimization, we define a potential dose range of between 0 and 200 mg (`mina` and `maxa`), and a range of potential sample times between 0 and 10 hours for the first three samples and between 240 and 248 hours for the last two samples (`minxt` and `maxxt`).   

Finally, we fix the two groups of subjects to have the sample sample times (`bUseGrouped_xt=TRUE`). 

```{r}
poped.db <- create.poped.database(ff_fun=ff,
                                  fg_fun=sfg,
                                  fError_fun=feps,
                                  bpop=c(V=72.8,KA=0.25,CL=3.75,Favail=0.9), 
                                  notfixed_bpop=c(1,1,1,0),
                                  d=c(V=0.09,KA=0.09,CL=0.25^2), 
                                  sigma=c(0.04,5e-6),
                                  notfixed_sigma=c(1,0),
                                  m=2,
                                  groupsize=20,
                                  a=list(c(DOSE=20,TAU=24),c(DOSE=40, TAU=24)),
                                  maxa=c(DOSE=200,TAU=24),
                                  mina=c(DOSE=0,TAU=24),
                                  xt=c( 1,2,8,240,245),
                                  minxt=c(0,0,0,240,240),
                                  maxxt=c(10,10,10,248,248),
                                  bUseGrouped_xt=TRUE)
```

##  The tasks to perform.

### Simulation
First it may make sense to check your model and design to make sure you get what you expect when simulating data.  Here we plot the model typical values:
```{r, fig.width=6}
plot_model_prediction(poped.db, model_num_points = 500)
```

Next, we plot the model typical values prediction intervals taking into account the between-subject variability (you can even investigate the effects of the residual, unexplained, variability with the `DV=T` argument): 
```{r simulate_with_BSV, fig.width=6}
plot_model_prediction(poped.db, model_num_points=500, IPRED=T)
```

We can get these predictions numerically as well:
```{r}
dat <- model_prediction(poped.db,DV=T)
head(dat,n=5);tail(dat,n=5)
```

### Design evaluation
Next, we evaluate the initial design
```{r}
evaluate_design(poped.db)
```
We see that the fixed-effect and residual variability parameters are relatively well estimated with this design, but the between-subject variability parameters are less well estimated.

### Evaluate alternative design
We can also compare if it is better to have more samples per subject or more subjects in the study - keeping the total number of observations fixed (5 samples/subjects * 20 subjects/group = 100 observations).
We now change the group size to 25 and sample at 4 time points: xt = c(1,6,240,245).

```{r,include = FALSE}
poped.db.new <- create.poped.database(ff_fun=ff,
                                  fg_fun=sfg,
                                  fError_fun=feps,
                                  bpop=c(V=72.8,KA=0.25,CL=3.75,Favail=0.9), 
                                  notfixed_bpop=c(1,1,1,0),
                                  d=c(V=0.09,KA=0.09,CL=0.25^2), 
                                  sigma=c(0.04,5e-6),
                                  notfixed_sigma=c(1,0),
                                  m=2,
                                  groupsize=25,
                                  a=list(c(DOSE=20,TAU=24),c(DOSE=40, TAU=24)),
                                  maxa=c(DOSE=200,TAU=24),
                                  mina=c(DOSE=0,TAU=24),
                                  xt=c( 1,6,240,245),
                                  minxt=c(0,0,240,240),
                                  maxxt=c(10,10,248,248),
                                  bUseGrouped_xt=TRUE)
```
```{r}
evaluate_design(poped.db.new)
```
### Comparison of designs
As you can see the precision on CL (bpop[3]) is better with the alternative design but the other parameters are less well estimated. So which of these designs is better?

We can compare the objective function value (ofv) and see that for this dummy example the alternative design (more subjects per group, less samples per subject) the ofv is higher (=better).
This value can also be used to optimize sampling times, doses, and other design features.

### Design optimization
Now we can optimize the sample times of the design^[Tip:  to make the optimization run faster use the option `parallel = T` in the `poped_optim` command.].
```{r optimize,message = FALSE,results='hide',cache=TRUE}
output <- poped_optim(poped.db, opt_xt=T)
```


```{r, fig.width=6}
summary(output)
plot_model_prediction(output$poped.db)
```


We see that there are four distinct sample times for this design.  This means that for this model, with these exact parameter values, that the most information from the study to inform the parameter estimation is with these sample times.  

### Examine efficiency of sampling windows
Of course, this means that there are multiple samples at some of these time points.  We can explore a more practical design by looking at the loss of efficiency if we spread out sample times in a uniform distribution around these optimal points ($\pm 30$ minutes).  
```{r,fig.width=6,fig.height=6,cache=TRUE}
plot_efficiency_of_windows(output$poped.db,xt_windows=0.5)
```

Here we see the efficiency ($(|FIM_{optimized}|/|FIM_{initial}|)^{1/npar}$) drop below 80% in some cases, which is mostly caused by an increase in the D[2,2] parameter uncertainty (BSV on absorption).  Smaller windows or different windowing on different samples may be in order here.  To investigate see `?plot_efficiency_of_windows`. 

### Optimize over a discrete design space
In the previous example we optimized over a continuous design space (sample times could be optimized to be any value between a lower and an upper limit).  We could also limit the search to only "allowed" values, for example, only samples taken on the hour are allowed.

```{r, message = FALSE,results='hide'}
poped.db.discrete <- create.poped.database(poped.db,discrete_xt = list(0:248))
                                          
output_discrete <- poped_optim(poped.db.discrete, opt_xt=T)

```

```{r,fig.width=6}
summary(output_discrete)
plot_model_prediction(output_discrete$poped.db)
```

Here we see that the optimization ran somewhat quicker, but gave a less efficient design.

### Optimize 'Other' design variables
One could also optimize over dose, to see if a different dose could help in parameter estimation .
```{r optimize_dose,message = FALSE,results='hide', eval=FALSE,cache=TRUE}
output_dose_opt <- poped_optim(output$poped.db, opt_xt=T, opt_a=T)
```


In this case the results are predictable ... higher doses give observations with somewhat lower absolute residual variability leading to both groups at the highest allowed dose levels (200 mg in this case).

### Cost function to optimize dose
Optimizing the dose of a study just to have better model parameter estimates may be somewhat implausible. Instead, let's use a cost function to optimize dose based on some sort of target concentration ... perhaps typical population trough concentrations of 0.2 and 0.35 for the two groups of patients at 240 hours. 

First we define the criteria we use to optimize the doses, here a least squares minimization.
```{r}
crit_fcn <- function(poped.db,...){
  pred_df <- model_prediction(poped.db)
  sum((pred_df[pred_df["Time"]==240,"PRED"] - c(0.2,0.35))^2)
}
crit_fcn(output$poped.db)
```

Now we minimize the cost function 
```{r cost_optimization, message = FALSE,results='hide',cache=TRUE}
output_cost <- poped_optim(poped.db, opt_a =T, opt_xt = F,
                     ofv_fun=crit_fcn,
                     maximize = F)
```

```{r, fig.width=6}
summary(output_cost)
get_rse(output_cost$FIM, output_cost$poped.db)
plot_model_prediction(output_cost$poped.db)
```

# Examples

In this section, further PopED features are described within a series of examples that are stored at the usual directory, located at:

```
system.file("examples", package="PopED")
```

Of note, only the code related to the specific "new" feature is here described. The complete R code is in the corresponding R script for each of the examples.

The table below provides a check list of features for each of the available examples.

| Features  | Ex1   | Ex2 |Ex3|Ex4|Ex5|Ex6|Ex7|Ex8|Ex9|Ex10|Ex11|Ex12|Ex13|Ex14|
|  -----   |  ----   | ----  |---- |---- |---- |---- |---- |---- |---- |---- |---- |---- |---- |---- |
| Library of models   |    |  |||||||||||||
| User defined models  |    |  |||||||||||||
| Multi response models  |    |  |||||||||||||
| Discrete covariates   |    |  |||||||||||||
| Continuous covariates |    |  |||||||||||||
| Full cov matrix W |    |  |||||||||||||
| Full cov matrix Sigma |    |  |||||||||||||
| Inter-occ variability  |    |  |||||||||||||
| Handling BLQ data  |    |  |||||||||||||
| Handling multiple <br> doses for ODEs  |    |  |||||||||||||
| Different dose <br> regimens per arm  |    |  |||||||||||||
| Irregular dosing <br> (e.g. dose weekly, <br> later monthly)  |    |  |||||||||||||
| Irregular dosing <br> (e.g. dose weekly, <br> later monthly)  |    |  |||||||||||||
| Irregular dosing <br> (e.g. more admin type <br> within one arm)  |    |  |||||||||||||
| Symbolic <br> differentiation  |    |  |||||||||||||
| Numeric <br> differentiation  |    |  |||||||||||||
| Automatic <br> differentiation |    |  |||||||||||||
| Non-Block FIM  |    |  |||||||||||||
| Designs differ <br> across responses   |    |  |||||||||||||
| Bayesian FIM    |    |  |||||||||||||
| First order <br> approximation (FO)   |    |  |||||||||||||
| Discrete ( e.g. optimizing among <br> a certain number of <br> discrete times)   |    |  |||||||||||||
| Evaluation/Optimization <br> of more-group designs<br> (same response)   |    |  |||||||||||||
| Different optimal<br> sampling times<br> between groups   |    |  |||||||||||||
| Constraining the <br>optimization to different <br> allowed sampling times<br> for each group   |    |  |||||||||||||
| Constraining the <br>optimization to different <br> allowed sampling times<br> for each response   |    |  |||||||||||||
| Keep some sampling time<br> fixed (they will be<br> automatically part of the <br>optimal design protocol)   |    |  |||||||||||||
| Optimization of subjects <br>per group    |    |  |||||||||||||
| Simultaneous <br>optimization of dose <br>and sampling times   |    |  |||||||||||||
| Handling <br>derived outputs   |    |  |||||||||||||
| Calculate precision <br>of derived parameters   |    |  |||||||||||||
| Power calculation   |    |  |||||||||||||
| Include previous FIM   |    |  |||||||||||||
| Compute shrinkage   |    |  |||||||||||||
| Fedorov-Wynn ("FW")  |    |  |||||||||||||
| Simplex ("SIMP")  |    |  |||||||||||||
| "D-optimality". <br>Determinant of <br>the FIM: det(FIM)  |    |  |||||||||||||


##1.	One compartment first order absorption - analytical solution for both mutiple and single dosing (design evaluation); design evaluation and design optimization of times and/or doses over a continuous or discrete design space

This example is the intro example illustrated in the first part of this document, before the Examples section.
The corresponding R script is the PopED example "ex.1.a.PK.1.comp.oral.md.intro.R"

Evaluation/Optimization of more-group designs (same response)

Different optimal sampling times between groups

##2.	PKPD model, multiple arms for different doses and sampling times, using analytical equations 

The full code for this example is available at the usual PopED examples directory, under the name of "ex.4.PKPD.1.comp.emax.R". It would be nice to have the plot function "plot_model_prediction" updated for this example, so that in the plot we see the name of the models, i.e. PK and PD, instead of the current predefined form, i.e. Model 1 and Model 2.

### Define the PKPD model

Here we define a PKPD mode using analytical equations. The PK is a one compartment model with intravenus bolus administration and linear elimination, whereas the PD is an ordinary Emax model.
```{r}
f_pkpdmodel <- function(model_switch,xt,parameters,poped.db){
  with(as.list(parameters),{
    y=xt
    MS <- model_switch
    
    # PK model
    CONC = DOSE/V*exp(-CL/V*xt) 
    
    # PD model
    EFF = E0 + CONC*EMAX/(EC50 + CONC)
    
    y[MS==1] = CONC[MS==1]
    y[MS==2] = EFF[MS==2]
    
    return(list( y= y,poped.db=poped.db))
  })
}
```

```{r, echo=F}
## -- parameter definition function
## ETAs into Parameter Values
f_etaToParam <- function(x,a,bpop,b,bocc){
  parameters=c( 
    CL=bpop[1]*exp(b[1])  ,
    V=bpop[2]*exp(b[2])	,
    E0=bpop[3]*exp(b[3])	,
    EMAX=bpop[4],
    EC50=bpop[5]*exp(b[4])	,
    DOSE=a[1]
  )
  return( parameters ) 
}
```

The error model has to accommodate both response models:
```{r}
## -- Residual Error function
## -- Proportional PK + additive PD
f_Err <- function(model_switch,xt,parameters,epsi,poped.db){
  returnArgs <- do.call(poped.db$model$ff_pointer,list(model_switch,xt,parameters,poped.db)) 
  y <- returnArgs[[1]]
  poped.db <- returnArgs[[2]]
  
  MS <- model_switch
  
  prop.err <- y*(1+epsi[,1])
  add.err <- y+epsi[,2]
  
  y[MS==1] = prop.err[MS==1]
  y[MS==2] = add.err[MS==2]
  
  return(list( y= y,poped.db =poped.db )) 
}
```

### Define design and design space

We specify in the poped.db object the vector "model_switch" in order to assign the sampling times written in the vector "xt" to the PK (1) or PD (2) models. 

```{r}
poped.db <- create.poped.database(ff_fun=f_pkpdmodel,
                                  fError_fun=f_Err,
                                  fg_fun=f_etaToParam,
                                  groupsize=20,
                                  m=3,
                                  sigma=diag(c(0.15,0.015)),
                                  bpop=c(CL=0.5,V=0.2,E0=1,EMAX=1,EC50=1),  
                                  d=c(CL=0.09,V=0.09,E0=0.04,EC50=0.09), 
                                  xt=c( 0.33,0.66,0.9,5,0.1,1,2,5),
                                  bUseGrouped_xt=1,
                                  model_switch=c(1,1,1,1,2,2,2,2),
                                  minxt=0,
                                  maxxt=5,
                                  ourzero = 0,
                                  a=list(c(DOSE=0),c(DOSE=1),c(DOSE=2)),
                                  maxa=c(DOSE=10),
                                  mina=c(DOSE=0))


```

Below the model predictions of the typical PK and PD profiles, i.e. Model 1 and Model 2, respectively, for three dose groups, are displayed. The current design, as shown in the "poped.db" object, consists in 3 arms with dose 0,1, and 2 (mg?) per arm; PK times of 0.33,0.66,0.9,5 (hours ?); PD times of 0.1,1,2,5 (hours?).

```{r,fig.width=6,fig.height=6}
plot_model_prediction(poped.db,IPRED=T,DV=T,facet_scales="free",separate.groups=T) 
```

##3.	Two compartment model with oral absorption using ODEs. Multiple dose dosing scheme.

## TODO: Martin please add rcpp code to this example.

The full code for this example is available at the usual PopED examples directory, under the name of "ex.9.PK.2.comp.oral.md.ode.compiled.R". 

Of note, the "deSolve" library needs to be installed for computing the system of differential equations. 

### Define the two compartment model

```{r}
#' Define the ODE system
PK.2.comp.oral.ode <- function(Time, State, Pars){
  with(as.list(c(State, Pars)), {    
    dA1 <- -KA*A1 
    dA2 <- KA*A1 + A3* Q/V2 -A2*(CL/V1+Q/V1)
    dA3 <- A2* Q/V1-A3* Q/V2
    return(list(c(dA1, dA2, dA3)))
  })
}
```

Define the initial conditions and the dosing

```{r}
ff.PK.2.comp.oral.md.ode <- function(model_switch, xt, parameters, poped.db){
  with(as.list(parameters),{
    A_ini <- c(A1=0, A2=0, A3=0)
    times_xt <- drop(xt)
    dose_times = seq(from=0,to=max(times_xt),by=TAU)
    eventdat <- data.frame(var = c("A1"), 
                           time = dose_times,
                           value = c(DOSE), method = c("add"))
    times <- sort(c(times_xt,dose_times))
    out <- ode(A_ini, times, PK.2.comp.oral.ode, parameters,
           events = list(data = eventdat))#atol=1e-13,rtol=1e-13)
    y = out[, "A2"]/(V1/Favail)
    y=y[match(times_xt,out[,"time"])]
    y=cbind(y)
    return(list(y=y,poped.db=poped.db))
  })
}

```


```{r, echo=F}
#' parameter definition function 
#' names match parameters in function ff
fg <- function(x,a,bpop,b,bocc){
  parameters=c( CL=bpop[1]*exp(b[1]),
                V1=bpop[2],
                KA=bpop[3]*exp(b[2]),
                Q=bpop[4],
                V2=bpop[5],
                Favail=bpop[6],
                DOSE=a[1],
                TAU=a[2])
  return( parameters ) 
}
```

### Define design and design space

```{r}
#' create poped database

discrete_a <- cell(2,1)

poped.db <- create.poped.database(ff_fun="ff.PK.2.comp.oral.md.ode",
                                  fError_fun="feps.add.prop",
                                  fg_fun="fg",
                                  groupsize=20,
                                  m=1,      #number of groups
                                  sigma=c(prop=0.1^2,add=0.05^2),
                                  bpop=c(CL=10,V1=100,KA=1,Q= 3.0, V2= 40.0, Favail=1),
                                  d=c(CL=0.15^2,KA=0.25^2),
                                  notfixed_bpop=c(1,1,1,1,1,0),
                                  xt=c( 48,50,55,65,70,85,90,120), 
                                  minxt=0,
                                  maxxt=144,
                                  discrete_xt = list(0:144),
                                  a=c(DOSE=100,TAU=24),
                                  maxa=c(DOSE=1000,TAU=24),
                                  mina=c(DOSE=0,TAU=8),
                                  discrete_a = list(DOSE=seq(0,1000,by=100),TAU=8:24))
```

### Plot intial design (just PRED)

```{r,fig.width=6}
plot_model_prediction(poped.db)
```

### Design evaluation

```{r}
#' how long does one evaluation of the FIM take? 
#tic(); (eval <- evaluate_design(poped.db)); toc()
eval <- evaluate_design(poped.db)
eval$ofv
eval$rse
```

TODO: To make optimization more reasonable you can use compiled code...
 




##4.	Different dose regimens per arm	

The full code for this example is available at the usual PopED examples directory, under the name of "ex.8.tmdd_qss_one_target_compiled.R".


```{r, echo=F, results="hide"}
### Compiled code for tmdd model

#We write the model using compiled code according to the Rcpp package.

cppFunction('List tmdd_qss_one_target_model_ode
            (double Time, NumericVector A, NumericVector Pars) {
            int n = A.size();
            NumericVector dA(n);

            double CL = Pars[0];
            double V1 = Pars[1];
            double Q  = Pars[2];
            double V2 = Pars[3];
            double FAVAIL = Pars[4];
            double KA = Pars[5];
            double VMAX = Pars[6];
            double KMSS = Pars[7];
            double R0 = Pars[8];
            double KSSS = Pars[9];
            double KDEG = Pars[10];
            double KINT = Pars[11];
            double DOSE = Pars[12];
            double SC_FLAG = Pars[13];

            double RTOT, CTOT, CFREE;
            RTOT = A[3];
            CTOT= A[1]/V1;
            CFREE = 0.5*((CTOT-RTOT-KSSS)+sqrt((CTOT-RTOT-KSSS)*(CTOT-RTOT-KSSS)+4*KSSS*CTOT));

            dA[0] = -KA*A[0];
            dA[1] = FAVAIL*KA*A[0]+(Q/V2)*A[2]-
                    (CL/V1+Q/V1)*CFREE*V1-RTOT*KINT*CFREE*V1/(KSSS+CFREE);
            dA[2] = (Q/V1)*CFREE*V1 - (Q/V2)*A[2];
            dA[3] = R0*KDEG - KDEG*RTOT - (KINT-KDEG)*(RTOT*CFREE/(KSSS+CFREE));
            return List::create(dA);
}')
```


```{r,echo=F, results="hide"}
## Define parameters
sfg <- function(x,a,bpop,b,bocc){
  parameters=c( CL=bpop[1]*exp(b[1])  ,
                V1=bpop[2]*exp(b[2])	,
                Q=bpop[3]*exp(b[3])	,
                V2=bpop[4]*exp(b[4])	,
                FAVAIL=bpop[5]*exp(b[5])	,
                KA=bpop[6]*exp(b[6])	,                       
                VMAX=bpop[7]*exp(b[7])	,
                KMSS=bpop[8]*exp(b[8])	,
                R0=bpop[9]*exp(b[9])	,
                KSSS=bpop[10]*exp(b[10])	,
                KDEG=bpop[11]*exp(b[11])	,
                KINT=bpop[12]*exp(b[12])	,
                DOSE=a[1]	,
                SC_FLAG=a[2])   
  return(parameters) 
}
```

```{r,echo=F}
tmdd_qss_one_target_model_compiled <- function(model_switch,xt,parameters,poped.db){
  with(as.list(parameters),{
    y=xt

    #The initialization vector for the compartment
    A_ini <- c(A1=DOSE*SC_FLAG,
               A2=DOSE*(1-SC_FLAG),
               A3=0,
               A4=R0)

    #Set up time points for the ODE
    times_xt <- drop(xt)
    times <- sort(times_xt)
    times <- c(0,times) ## add extra time for start of integration

    # solve the ODE
    out <- ode(A_ini, times, tmdd_qss_one_target_model_ode, parameters)#,atol=1e-13,rtol=1e-13)


    # extract the time points of the observations
    out = out[match(times_xt,out[,"time"]),]

    # Match ODE output to measurements
    RTOT = out[,"A4"]
    CTOT = out[,"A2"]/V1
    CFREE = 0.5*((CTOT-RTOT-KSSS)+sqrt((CTOT-RTOT-KSSS)^2+4*KSSS*CTOT))
    COMPLEX=((RTOT*CFREE)/(KSSS+CFREE))
    RFREE= RTOT-COMPLEX

    y[model_switch==1]= RTOT[model_switch==1]
    y[model_switch==2] =CFREE[model_switch==2]
    #y[model_switch==3]=RFREE[model_switch==3]

    return(list( y=y,poped.db=poped.db))
  })
}

```

```{r, echo=F}
tmdd_qss_one_target_model_ruv <- function(model_switch,xt,parameters,epsi,poped.db){
  returnArgs <- do.call(poped.db$model$ff_pointer,list(model_switch,xt,parameters,poped.db))
  y <- returnArgs[[1]]
  poped.db <- returnArgs[[2]]

  y[model_switch==1] = log(y[model_switch==1])+epsi[,1]
  y[model_switch==2] = log(y[model_switch==2])+epsi[,2]

  return(list(y=y,poped.db=poped.db))
}

```

### Define the design 

Two different studies are defined, with different sampling times in terms of total number and values.
For this reason it is important to specify the number of samples per arm by defining "ni" in the
create.poped.database() function. It is also necessary to define the matrix "G_xt", as shown below. 

```{r}
#################################################
# for study 1 + 2 in gibiansky,JPKPD,2012 table 2 
#################################################

xt <- zeros(6,30)
study_1_xt <- matrix(rep(c(0.0417,0.25,0.5,1,3,7,14,21,28,35,42,49,56),8),nrow=4,byrow=T)
study_2_xt <- matrix(rep(c(0.0417,1,1,7,14,21,28,56,63,70,77,84,91,98,105),4),nrow=2,byrow=T)
xt[1:4,1:26] <- study_1_xt
xt[5:6,] <- study_2_xt

model_switch <- zeros(6,30)
model_switch[1:4,1:13] <- 1
model_switch[1:4,14:26] <- 2
model_switch[5:6,1:15] <- 1
model_switch[5:6,16:30] <- 2

G_xt <- zeros(6,30)
study_1_G_xt <- matrix(rep(c(1:13),8),nrow=4,byrow=T)
study_2_G_xt <- matrix(rep(c(14:28),4),nrow=2,byrow=T)
G_xt[1:4,1:26] <- study_1_G_xt
G_xt[5:6,] <- study_2_G_xt

poped.db.2 <- create.poped.database(ff_fun=tmdd_qss_one_target_model_compiled,
                                    fError_fun=tmdd_qss_one_target_model_ruv,
                                    fg_fun=sfg,
                                    groupsize=rbind(6,6,6,6,100,100),
                                    m=6,      #number of groups
                                    sigma=c(0.04,0.0225),
                                    bpop=c(CL=0.3,V1=3,Q=0.2,V2=3,FAVAIL=0.7,KA=0.5,VMAX=0,
                                           KMSS=0,R0=0.1,KSSS=0.015,KDEG=10,KINT=0.05),
                                    d=c(CL=0.09,V1=0.09,Q=0.04,V2=0.04,FAVAIL=0.04,
                                        KA=0.16,VMAX=0,KMSS=0,R0=0.09,KSSS=0.09,KDEG=0.04,
                                        KINT=0.04),
                                    notfixed_bpop=c( 1,1,1,1,1,1,0,0,1,1,1,1),
                                    notfixed_d=c( 1,1,1,1,1,1,0,0,1,1,1,1),
                                    xt=xt,
                                    model_switch=model_switch,
                                    ni=rbind(26,26,26,26,30,30),
                                    bUseGrouped_xt=1,
                                    G_xt=G_xt,
                                    a=list(c(DOSE=100, SC_FLAG=0),
                                           c(DOSE=300, SC_FLAG=0),
                                           c(DOSE=600, SC_FLAG=0),
                                           c(DOSE=1000, SC_FLAG=1),
                                           c(DOSE=600, SC_FLAG=0),
                                           c(DOSE=1000, SC_FLAG=1)),
                                    discrete_a = list(DOSE=seq(100,1000,by=100),
                                                      SC_FLAG=c(0,1)))

```

### Plot model predisctions
```{r, fig.width=6}
plot_model_prediction(poped.db.2,facet_scales="free")
```

### Design evaluation

```{r}
tic(); eval_2 <- evaluate_design(poped.db.2); toc()
eval_2$ofv
eval_2$rse
```

##5.	Model with discrete covariates	
## TODO --> Martin

##6.	Model with continuous covariates

The R code for this example is available at the usual PopED examples directory, under the name of "covariate_distributions.R"

TODO: reduce example, if needed.

## Introduction
Lets assume that we have a model with a covariate included in the model description.  Here we define a one-compartment PK model that has weight on both clearance and volume of distribution. 

```{r model}
mod_1 <- function(model_switch,xt,parameters,poped.db){
  with(as.list(parameters),{
    y=xt
    
    CL=CL*(WT/70)^(WT_CL)
    V=V*(WT/70)^(WT_V)
    DOSE=1000*(WT/70)
    y = DOSE/V*exp(-CL/V*xt) 
    
    return(list( y= y,poped.db=poped.db))
  })
}

par_1 <- function(x,a,bpop,b,bocc){
  parameters=c( CL=bpop[1]*exp(b[1]),
                V=bpop[2]*exp(b[2]),
                WT_CL=bpop[3],
                WT_V=bpop[4],
                WT=a[1])
  return( parameters ) 
}
```

Now we define a design.  In this case one group of individuals, where we define the individuals' typical weight as 70 kg. 
```{r design}
poped_db <- create.poped.database(ff_fun=mod_1,
                                  fg_fun=par_1,
                                  fError_fun=feps.add.prop,
                                  groupsize=50,
                                  m=1,
                                  sigma=c(0.015,0.0015),
                                  notfixed_sigma = c(1,0),
                                  bpop=c(CL=3.8,V=20,WT_CL=0.75,WT_V=1), 
                                  d=c(CL=0.05,V=0.05), 
                                  xt=c( 1,2,4,6,8,24),
                                  minxt=0,
                                  maxxt=24,
                                  bUseGrouped_xt=1,
                                  a=c(WT=70)
                                  )
```

We can create a plot of the model typical predictions:
```{r}
plot_model_prediction(poped_db)
```


And evaluate the initial design
```{r}
evaluate_design(poped_db)
```

We see that the covariate parameters can not be estimated according to this design calculation (RSE of bpop[3]=0 and bpop[4]=0).  Why is that? Well, the calculation being done is assuming that every individual in the group has the same covariate (to speed up the calculation).  This is clearly a poor prediction in this case!

## Distribution of covariates

We can improve the computation by assuming a distribution of the covariate (WT) in the individuals in the study. We set `groupsize=1`, the number of groups to be 50 (`m=50`) and assume that WT is sampled from a normal distribution with mean=70 and sd=10 (`a=as.list(rnorm(50,mean = 70,sd=10)`).
```{r}
poped_db_2 <- create.poped.database(ff_fun=mod_1,
                                  fg_fun=par_1,
                                  fError_fun=feps.add.prop,
                                  groupsize=1,
                                  m=50,
                                  sigma=c(0.015,0.0015),
                                  notfixed_sigma = c(1,0),
                                  bpop=c(CL=3.8,V=20,WT_CL=0.75,WT_V=1), 
                                  d=c(CL=0.05,V=0.05), 
                                  xt=c( 1,2,4,6,8,24),
                                  minxt=0,
                                  maxxt=24,
                                  bUseGrouped_xt=1,
                                  a=as.list(rnorm(50,mean = 70,sd=10))
                                  )

```

```{r}
ev <- evaluate_design(poped_db_2) 
ev$ofv
ev$rse
```

Here we see that, given this distribution of weights, the covariate effect parameters (bpop[3] and bpop[4]) would be well estimated.

However, we are only looking at one sample of 50 individuals.  Maybe a better approach is to look at the distribution of RSEs over a number of experiments given the expected weight distribution. 

```{r}
nsim <- 10
rse_list <- c()
for(i in 1:nsim){
  poped_db_tmp <- create.poped.database(ff_fun=mod_1,
                                        fg_fun=par_1,
                                        fError_fun=feps.add.prop,
                                        groupsize=1,
                                        m=50,
                                        sigma=c(0.015,0.0015),
                                        notfixed_sigma = c(1,0),
                                        bpop=c(CL=3.8,V=20,WT_CL=0.75,WT_V=1), 
                                        d=c(CL=0.05,V=0.05), 
                                        xt=c( 1,2,4,6,8,24),
                                        minxt=0,
                                        maxxt=24,
                                        bUseGrouped_xt=1,
                                        a=as.list(rnorm(50,mean = 70,sd=10)))
  rse_tmp <- evaluate_design(poped_db_tmp)$rse
  rse_list <- rbind(rse_list,rse_tmp)
}
apply(rse_list,2,quantile)
```

##7.	Model with Inter-occasion variability

The full code for this example is available at the usual PopED examples directory, under the name of "model_with_iov.R"

```{r}
##-- Model: One comp first order absorption
## -- Analytic solution for both mutiple and single dosing
ff <- function(model_switch,xt,parameters,poped.db){
  with(as.list(parameters),{
    y=xt
    N = floor(xt/TAU)+1
    CL = (xt*0+1)*CL_OCC_1
    CL[N>9] = CL_OCC_2
    y=(DOSE*Favail/V)*(KA/(KA - CL/V)) *
      (exp(-CL/V * (xt - (N - 1) * TAU)) * (1 - exp(-N * CL/V * TAU))/(1 - exp(-CL/V * TAU)) -
         exp(-KA * (xt - (N - 1) * TAU)) * (1 - exp(-N * KA * TAU))/(1 - exp(-KA * TAU)))
    return(list( y=y,poped.db=poped.db))
  })
}
```

The IOV is introduced with "bocc" in the parameter definition function sfg().

```{r, echo= FALSE}
## -- parameter definition function
## -- names match parameters in function ff
sfg <- function(x,a,bpop,b,bocc){
  parameters=c( V=bpop[1]*exp(b[1]),
                KA=bpop[2]*exp(b[2]),
                CL_OCC_1=bpop[3]*exp(b[3]+bocc[1,1]),
                CL_OCC_2=bpop[3]*exp(b[3]+bocc[1,2]),
                Favail=bpop[4],
                DOSE=a[1],
                TAU=a[2])
  return( parameters ) 
}

```

The within-subject variability variances (`docc`) are defined in poped.db.

```{r} 
## -- Define design and design space
poped.db <- create.poped.database(ff_fun=ff,
                                  fg_fun=sfg,
                                  fError_fun=feps.add.prop,
                                  bpop=c(V=72.8,KA=0.25,CL=3.75,Favail=0.9), 
                                  notfixed_bpop=c(1,1,1,0),
                                  d=c(V=0.09,KA=0.09,CL=0.25^2), 
                                  sigma=c(0.04,5e-6),
                                  notfixed_sigma=c(0,0),
                                  docc = matrix(c(0,0.09,0),nrow = 1),
                                  m=2,
                                  groupsize=20,
                                  xt=c( 1,2,8,240,245),
                                  minxt=c(0,0,0,240,240),
                                  maxxt=c(10,10,10,248,248),
                                  bUseGrouped_xt=1,
                                  a=list(c(DOSE=20,TAU=24),c(DOSE=40, TAU=24)),
                                  maxa=c(DOSE=200,TAU=24),
                                  mina=c(DOSE=0,TAU=24))
```

###Create plot of model without variability 

```{r,fig.width=6}
plot_model_prediction(poped.db, model_num_points = 300)
```

###Create plot of model with variability 

```{r,fig.width=6}
plot_model_prediction(poped.db, IPRED=T, DV=T, separate.groups=T, model_num_points = 300)
```

###Evaluate initial design

```{r}
ev <- evaluate_design(poped.db)
ev$ofv
ev$rse
```

##8.	Full omega matrix	

The full code for this example is available at the usual PopED examples directory, under the name of "example_convd_2.R".

```{r, echo=F, results="hide"}
ff <- function(model_switch,xt,parameters,poped.db){
  ##-- Model: One comp first order absorption
  with(as.list(parameters),{
    y=xt
    y=(DOSE*Favail*KA/(V*(KA-CL/V)))*(exp(-CL/V*xt)-exp(-KA*xt))
    return(list(y=y,poped.db=poped.db))
  })
}

sfg <- function(x,a,bpop,b,bocc){
  ## -- parameter definition function 
  parameters=c(CL=bpop[1]*exp(b[1]),
               V=bpop[2]*exp(b[2]),
               KA=bpop[3]*exp(b[3]),
               Favail=bpop[4],
               DOSE=a[1])
  return(parameters) 
}

feps <- function(model_switch,xt,parameters,epsi,poped.db){
  ## -- Residual Error function
  ## -- Proportional 
  returnArgs <- ff(model_switch,xt,parameters,poped.db) 
  y <- returnArgs[[1]]
  poped.db <- returnArgs[[2]]
  y = y*(1+epsi[,1])
  
  return(list(y=y,poped.db=poped.db)) 
}

## -- Define initial design  and design space
poped.db <- create.poped.database(ff_file="ff",
                                  fg_file="sfg",
                                  fError_file="feps",
                                  bpop=c(CL=0.15, V=8, KA=1.0, Favail=1), 
                                  notfixed_bpop=c(1,1,1,0),
                                  d=c(CL=0.07, V=0.02, KA=0.6), 
                                  sigma=0.01,
                                  groupsize=32,
                                  xt=c( 0.5,1,2,6,24,36,72,120),
                                  minxt=0,
                                  maxxt=120,
                                  a=70)
```

###Define initial design and design space with covariance 

The `covd` object is used for defining the covariances of the IIV variances.

```{r}
poped.db_with <- create.poped.database(ff_file="ff",
                                  fg_file="sfg",
                                  fError_file="feps",
                                  bpop=c(CL=0.15, V=8, KA=1.0, Favail=1), 
                                  notfixed_bpop=c(1,1,1,0),
                                  d=c(CL=0.07, V=0.02, KA=0.6), 
                                  covd = c(.03,.1,.09),
                                  sigma=0.01,
                                  groupsize=32,
                                  xt=c( 0.5,1,2,6,24,36,72,120),
                                  minxt=0,
                                  maxxt=120,
                                  a=70)
```

What do the covariances mean?

```{r}
(IIV <- poped.db_with$parameters$param.pt.val$d)
cov2cor(IIV)
```

###Create plot of model with variability 

```{r,fig.width=6}
library(ggplot2)
p1 <- plot_model_prediction(poped.db,IPRED=T)+ylim(0,12)
p2 <- plot_model_prediction(poped.db_with,IPRED=T) +ylim(0,12)
gridExtra::grid.arrange(p1, p2, nrow = 1)
```

###Evaluate initial design (no full covariance matrix)

```{r}
ev1 <- evaluate_design(poped.db)
ev2 <- evaluate_design(poped.db_with)

ev1$ofv
ev1$rse

ev2$ofv
ev2$rse
```

###Evaluate design with full FIM

```{r}
ev1 <- evaluate_design(poped.db, fim.calc.type=0)
ev2 <-evaluate_design(poped.db_with, fim.calc.type=0)

ev1$ofv
ev1$rse

ev2$ofv
ev2$rse
```

##9.	Include previous FIM

In this example you will learn how to incorporate prior knowledge to your current study.
First the FIM obtained from an experiment in adult is stored. Then this FIM is added to the current experiment in children.
The full code for this example is available at the usual PopED examples directory, under the name of "ex.11.PK.prior.R".

```{r}
##-- Model: One comp first order absorption
## -- Analytic solution for both mutiple and single dosing
ff <- function(model_switch,xt,parameters,poped.db){
  with(as.list(parameters),{
    y=xt
    N = floor(xt/TAU)+1
    y=(DOSE*Favail/V)*(KA/(KA - CL/V)) * 
      (exp(-CL/V * (xt - (N - 1) * TAU)) * (1 - exp(-N * CL/V * TAU))/(1 - exp(-CL/V * TAU)) - 
         exp(-KA * (xt - (N - 1) * TAU)) * (1 - exp(-N * KA * TAU))/(1 - exp(-KA * TAU)))  
    return(list( y=y,poped.db=poped.db))
  })
}
```

We define here the parameters for one compartment first order absorption model.
Of note, "isPediatric" is the covariate on clearance for pediatrics.

```{r, echo=F, results="hide"}
## -- parameter definition function 
## -- names match parameters in function ff
## -- note, covariate on clearance for pediatrics (using isPediatric x[1])
sfg <- function(x,a,bpop,b,bocc){
  parameters=c( V=bpop[1]*exp(b[1]),
                KA=bpop[2]*exp(b[2]),
                CL=bpop[3]*exp(b[3])*bpop[5]^x[1], # add covariate for pediatrics
                Favail=bpop[4],
                isPediatric = x[1],
                DOSE=a[1],
                TAU=a[2])
  return( parameters ) 
}
```

```{r, echo=F, results="hide"}
## -- Residual unexplained variablity (RUV) function
## -- Additive + Proportional  
feps <- function(model_switch,xt,parameters,epsi,poped.db){
  returnArgs <- do.call(poped.db$model$ff_pointer,list(model_switch,xt,parameters,poped.db)) 
  y <- returnArgs[[1]]
  poped.db <- returnArgs[[2]]
  
  y = y*(1+epsi[,1])+epsi[,2]
  
  return(list( y= y,poped.db =poped.db )) 
}
```

Define design and design space for adults

```{r}
## -- Define design and design space for adults (isPediatric = 0)
## Two arms, 5 time points
poped.db <- create.poped.database(ff_fun="ff",
                                  fg_fun="sfg",
                                  fError_fun="feps",
                                  bpop=c(V=72.8,KA=0.25,CL=3.75,Favail=0.9,pedCL=0.8), 
                                  notfixed_bpop=c(1,1,1,1,1),
                                  d=c(V=0.09,KA=0.09,CL=0.25^2), 
                                  sigma=c(0.04,5e-6),
                                  notfixed_sigma=c(0,0),
                                  m=2,
                                  groupsize=20,
                                  xt=c( 1,2,8,240,245),
                                  bUseGrouped_xt=1,
                                  a=list(c(DOSE=20,TAU=24),c(DOSE=40, TAU=24)),
                                  x=list(isPediatric = 0)
)
```

```{r,fig.width=6}
##  create plot of model without variability
plot_model_prediction(poped.db, model_num_points = 300)

```

```{r}
## To store FIM from adult design
## evaluate initial design
outAdult = evaluate_design(poped.db)
print(outAdult)
```

```{r}
## Define pediatric model/design
## One arm, 4 time points only - insufficient on its own
poped.db.ped <- create.poped.database(ff_fun="ff",
                                  fg_fun="sfg",
                                  fError_fun="feps",
                                  bpop=c(V=72.8,KA=0.25,CL=3.75,Favail=0.9,pedCL=0.8), 
                                  notfixed_bpop=c(1,1,1,1,1),
                                  d=c(V=0.09,KA=0.09,CL=0.25^2), 
                                  sigma=c(0.04,5e-6),
                                  notfixed_sigma=c(0,0),
                                  m=1,
                                  groupsize=6,
                                  xt=c( 1,2,6,240),
                                  bUseGrouped_xt=1,
                                  a=list(c(DOSE=40,TAU=24)),
                                  x=list(isPediatric = 1)
)
```


```{r,fig.width=6}
##  create plot of model without variability 
plot_model_prediction(poped.db.ped, model_num_points = 300)

## evaluate design of pediatrics only - insufficient
eval <- evaluate_design(poped.db.ped)
eval$ofv
eval$rse
```

```{r}
## Add adult prior
poped.db.all <- create.poped.database(
  poped.db.ped,
  prior_fim = outAdult$fim
)

## evaluate design using prior FIM from adults
eval <- evaluate_design(poped.db.all)
eval$ofv
eval$rse
```

##10.	Design evaluation including uncertainty in the model parameters (robust design)	

The aim is to evaluate a design incorporating uncertainty around parameter values in the model.
The full code for this example is available at the usual PopED examples directory, under the name of "ex.2.d.warfarin.ED.R".
This illustration is one of the Warfarin examples from software comparison in: Nyberg et al., "Methods and software tools for design evaluation for population pharmacokinetics-pharmacodynamics studies", Br. J. Clin. Pharm., 2014. 

### Define the model

```{r, echo=F, results="hide"}
sfg <- function(x,a,bpop,b,bocc){
  ## -- parameter definition function 
  parameters=c(CL=bpop[1]*exp(b[1]),
               V=bpop[2]*exp(b[2]),
               KA=bpop[3]*exp(b[3]),
               Favail=bpop[4],
               DOSE=a[1])
  return(parameters) 
}

ff <- function(model_switch,xt,parameters,poped.db){
  ##-- Model: One comp first order absorption
  with(as.list(parameters),{
    y=xt
    y=(DOSE*Favail*KA/(V*(KA-CL/V)))*(exp(-CL/V*xt)-exp(-KA*xt))
    return(list(y=y,poped.db=poped.db))
  })
}
```

Adding 10% uncertainty to all fixed effects (not Favail)

```{r}
bpop_vals <- c(CL=0.15, V=8, KA=1.0, Favail=1)
bpop_vals_ed <- cbind(ones(length(bpop_vals),1)*4, # log-normal distribution
                      bpop_vals,
                      ones(length(bpop_vals),1)*(bpop_vals*0.1)^2) # 10% of bpop value
bpop_vals_ed["Favail",] <- c(0,1,0)
bpop_vals_ed
```

### Define initial design  and design space

```{r}
poped.db <- create.poped.database(ff_fun=ff,
                                  fg_fun=sfg,
                                  fError_fun=feps.add.prop,
                                  bpop=bpop_vals_ed, 
                                  notfixed_bpop=c(1,1,1,0),
                                  d=c(CL=0.07, V=0.02, KA=0.6), 
                                  sigma=c(0.01,0.25),
                                  groupsize=32,
                                  xt=c( 0.5,1,2,6,24,36,72,120),
                                  minxt=0,
                                  maxxt=120,
                                  a=70,
                                  mina=0,
                                  maxa=100,
                                  ED_samp_size=20)

```

### Design evaluation

```{r}
tic(); output <- evaluate.e.ofv.fim(poped.db,ED_samp_size=20); toc()
output$E_ofv
output$E_fim
```



##11. Design evaluation for a subset of model parameters of interest (Ds optimality)

Ds-optimality is the criterion that can be used in case one is interested in estimating a subset (s) of the model parameters as precisely as possible (Atkinson 2007).
The full code for this example is available at the usual PopED examples directory, under the name of "ex.2.e.warfarin.Ds.R". This illustration is one of the Warfarin examples from software comparison in: Nyberg et al., "Methods and software tools for design evaluation for population pharmacokinetics-pharmacodynamics studies", Br. J. Clin. Pharm., 2014. 

### Define initial design  and design space

```{r, echo=FALSE}
sfg <- function(x,a,bpop,b,bocc){
  ## -- parameter definition function 
  parameters=c(CL=bpop[1]*exp(b[1]),
               V=bpop[2]*exp(b[2]),
               KA=bpop[3]*exp(b[3]),
               Favail=bpop[4],
               DOSE=a[1])
  return(parameters) 
}

ff <- function(model_switch,xt,parameters,poped.db){
  ##-- Model: One comp first order absorption
  with(as.list(parameters),{
    y=xt
    y=(DOSE*Favail*KA/(V*(KA-CL/V)))*(exp(-CL/V*xt)-exp(-KA*xt))
    return(list(y=y,poped.db=poped.db))
  })
}
```

For Ds optimality we add in the create.poped.database function the `ds_index` to indicate whether a paramter is interesting (=1) or not (=0). Moreover we need to set the `ofv_calc_type` to 6 for computing the Ds optimality criterion, as it is set to 4 by default, i.e. for computing the log of the determiant of FIM. More details are available in the help of R, by runiing the command `?create.poped.database`.

```{r}
poped.db <- create.poped.database(ff_fun=ff,
                                  fg_fun=sfg,
                                  fError_fun=feps.add.prop,
                                  bpop=c(CL=0.15, V=8, KA=1.0, Favail=1), 
                                  notfixed_bpop=c(1,1,1,0),
                                  d=c(CL=0.07, V=0.02, KA=0.6), 
                                  sigma=c(0.01,0.25),
                                  groupsize=32,
                                  xt=c( 0.5,1,2,6,24,36,72,120),
                                  minxt=0,
                                  maxxt=120,
                                  a=70,
                                  mina=0,
                                  maxa=100,
                                  ds_index=c(0,0,0,1,1,1,1,1), # size is 
                                  #number_of_non_fixed_parameters
                                  ofv_calc_type=6) # Ds OFV calculation
```

### Design evaluation

```{r}
evaluate_design(poped.db)
```
##12.	Shrinkage	

The full code for this example is available at the usual PopED examples directory, under the name of "shrinkage.R".

```{r, echo=F, results="hide"}

  ## following example 1.4.2 in PFIM
  sfg <- function(x,a,bpop,b,bocc){
    ## -- parameter definition function 
    parameters=c(KA=bpop[1]*exp(b[1]),
                 K=bpop[2]*exp(b[2]),
                 V=bpop[3]*exp(b[3]),
                 DOSE=a[1])
    return(parameters) 
  }
  
  ff <- function(model_switch,xt,parameters,poped.db){
    ##-- Model: One comp first order absorption
    with(as.list(parameters),{
      y<-(DOSE/V*KA/(KA-K)*(exp(-K*xt)-exp(-KA*xt)))
      return(list(y=y,poped.db=poped.db))
    })
  }
  
  ## -- Residual unexplained variablity (RUV) function
  ## -- Additive + Proportional  
  feps <- function(model_switch,xt,parameters,epsi,poped.db){
    returnArgs <- do.call(poped.db$model$ff_pointer,list(model_switch,xt,parameters,poped.db)) 
    f <- returnArgs[[1]]
    y = f + (0.5 + 0.15*f)*epsi[,1]
    
    return(list( y= y,poped.db =poped.db )) 
  }
```

Define design and use function shrinkage().

```{r} 
  ## -- Define initial design  and design space
  poped.db <- create.poped.database(ff_fun=ff,
                                    fg_fun=sfg,
                                    fError_fun =feps,
                                    bpop=c(KA=2, K=0.25, V=15), 
                                    d=c(KA=1, V=0.25,0.1), 
                                    sigma=c(1),
                                    notfixed_sigma = c(0),
                                    groupsize=500,
                                    xt=c( 1,3,8),
                                    minxt=0,
                                    maxxt=10,
                                    a=100)
  
  
  shr_out <- shrinkage(poped.db)
  shr_out
```

##13.	Power Calculation	--> Martin

##14.	Espresso design	--> Martin

##Further examples	

15.	Handling BLQ data	
16.	Irregular dosing more complex: e.g. switching between s.c. and i.v. within one arm.	
17.	Constraining the optimization to different allowed sampling times for each group	
18.	Constraining the optimization to different allowed sampling times for each response	
19.	Keep some sampling time fixed (they will be automatically part of the optimal design protocol)	
20.	Handling derived outputs	
21.	Symbolic differentiation	
